### 2021-07-21 ch/3-structure.tex
We're nearly there!

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
% The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.

>>>>

The absolutely essential elements for this method are the encoder (also referred to as `de-biaser')
that encodes both, samples from the training set and samples from the deployment set,
to a splittable representation;
the adversary that tries to identify, from one part of the split representation,
which dataset the given samples originated from;
and some kind of reconstruction loss
to ensure the splittable representation represents the input data well.
The following elements are in theory optional,
but are needed for the method to work with real-world data:
clustering and sampling to ensure the training batches are balanced in specific ways;
and subdivision of the batches into bags and the aggregation of the adversarial loss over the bags.
Roughly speaking, these elements strengthen the supervision signal for the invariance learning.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
% The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.
The following is a detailed listing of all the individual author contributions.

### 2021-07-21 paper3/paper3.tex
We're nearly there!

\subsection{On disentangled representations learning.}
\citet{locatello2019fairness} suggested that disentanglement in representation learning may be a useful property to remove algorithmic bias when subgroup information is not observed.
%
In order for disentangled representations to reduce algorithmic bias without the knowledge of subgroup label $s$, they have to assume that the class label $y$ and the subgroup label $s$ are independent, i.e. $y\perp s$.
%
Though, in many real-world tasks, the variable $s$ is correlated with the variable $y$, and therefore unsupervised methods are not suitable  \citep{jaiswal2018unsupervised,JaiWuAbdNat19}. 
%
Indeed, experiments in \citet{locatello2019fairness} were wholly done with procedurally generated synthetic datasets involving 2D and 3D shapes. 
%
Without some supervision or inductive bias, disentangled representation methods would not solve the issue of algorithmic fairness with invisible demographics \citep{locatello2019challenging}.

@@

@@

>>>>

\subsection{On disentangled representations learning.}%
\label{ssec:on-disentangled-representation-learning}
\citet{locatello2019fairness} suggested that disentanglement in representation learning may be a useful property to remove algorithmic bias when subgroup information is not observed.
%
In order for disentangled representations to reduce algorithmic bias without the knowledge of subgroup label $s$, they have to assume that the class label $y$ and the subgroup label $s$ are independent, i.e. $y\perp s$.
%
Though, in many real-world tasks, the variable $s$ is correlated with the variable $y$, and therefore unsupervised methods are not suitable  \citep{jaiswal2018unsupervised,JaiWuAbdNat19}. 
%
Indeed, experiments in \citet{locatello2019fairness} were wholly done with procedurally generated synthetic datasets involving 2D and 3D shapes. 
%
Without some supervision or inductive bias, disentangled representation methods would not solve the issue of algorithmic fairness with invisible demographics \citep{locatello2019challenging}. 
\citet{locatello2020weakly} suggest that that it is possible to learn disentangled representations
with contrasting pairs that share at least one of the underlying factors but differ in some.
Here, the difference between the pairs acts as the supervision signal.

@@

As mentioned in \secref{ssec:on-disentangled-representation-learning},
\citet{locatello2020weakly} learned to disentangle from non-i.i.d.\ pairs
which share at least one underlying factor.
In our scenario, constructing the pairs each out of one sample from the training set and one sample from the deployment set
would satisfy this requirement if we allow the `underlying factor' to be quite abstract
(see also the section below for a discussion of this).
However, constructing the pairs each out of two \emph{bags} instead of two samples
makes the analysis much clearer:
bags from the training and deployment set are constructed such
that they both have the same distribution of \(y\) values (a uniform distribution over \(y\)),
but they differ in the distribution of \(s\) values.

@@

Another take-away from our work concerns the general disentangling setup described in
\citet{locatello2020weakly}
-- \ie\ using pairs that share at least one underlying factor but also differ some.
For our work, we found it advantageous to do this pair matching not on the level of samples,
but on the level of bags (\ie, small, specially-sampled \emph{sets} of samples).
It is possible that other disentangling problems would benefit
from a similar switch in perspective
that identifies not the samples but some other structure
as the natural way to construct the pairs for the disentangling.
